Demes: a specification for interchanging complex demography

### 2022-03-15 paper.tex
begin the overhaul

@@

\abstract{
Understanding the demographic history of populations is a
key goal in population genetics, and with improving methods
and data, ever more complex models are being estimated.
Demographic models of current interest (consisting of a set of discrete populations,
their sizes and growth rates, and continuous and pulse migrations
between those populations over a number of epochs) can require
dozens of parameters to fully describe. There is currently
no standard format to define such models, significantly
hampering  progress in the field. In particular, the important
task of translating the model descriptions in published
work into input suitable for population genetic simulators is labour intensive
and error prone.
We propose the Demes data model and file format,
built on widely used technologies,
to alleviate these issues. Demes provides a well-defined and unambiguous
graphical model of populations and their properties that is straightforward to
implement in software, and a text file format that is designed for
simplicity and clarity.
We provide thoroughly tested implementations of Demes parsers in Python
and C, and showcase initial support in a number of simulators and inference
methods.
}

@@

The ever-increasing amount of genetic sequencing data from genetically and
geographically diverse species and populations has allowed us to infer complex
demography and study life history at fine scales. An integral component to such
population genetics studies is simulation. Software to either simulate whole
genome sequences
\citep{thornton2014cpp,thornton2019-nu,staab2015scrm,baumdicker2021-iu,kelleher2016efficient,haller2019slim}
or informative summary statistics of diversity
\citep{gutenkunst2009inferring,kamm2017efficient,jouganous2017inferring} have
enabled the increasing complexity of genomic studies, with several software
packages capable of working with large sample sizes, many interacting populations, and
deviations from random-mating models of panmixia.
This ability to infer and simulate such complex demographic scenarios, however,
has highlighted a major shortcoming in community standards. The fragmented
landscape of different approaches to describing demographic models makes
it difficult to compare inferences made by different methods,
and to reliably simulate from previously inferred models.

Inference methods for demography usually output a set of
text files which describe the inferred model,
and nearly all methods use different file formats.
% Thus, any comparison
% between inferences requires these file formats to be parsed, with the user
% being left with the task of unifying the data models.
Inference results are typically reported in publications
via a combination of visual depiction,
a list of key parameters in tabular form and a discussion within the text.
Unfortunately there is often ambiguity in these descriptions, and
implementing the precise model inferred for later simulation
is at best tedious and error prone~\citep{ragsdale2020lessons}
and occasionally impossible because of missing information.
The \stdpopsim\ project is a community-maintained catalog of species
information and demographic models designed to provide standardised
population genetic simulations~\citep{adrion2020community}.
The development process for adding demographic models to
\stdpopsim\ tries to avoid errors in model implementation by requiring two
independent implementations of each model
to be equivalent. While this process is
robust and has uncovered both significant
errors~\citep{ragsdale2020lessons} and ambiguities in
published models, it is labour intensive.
It would be far better if published results
were reported in an unambiguous standardised format that could be provided
directly as input to simulators.

Simulation is a core tool in population genetics, and a
large number of different methods have been developed
over the past three decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer}.
Simulations are based on highly idealised population models
and one of the key uses of inferred demographic histories
is to make simulations more realistic.
Simulation methods take three broad approaches to specifying
the demographic model to simulate,
using either a command line
interface~\citep[e.g.][]{hudson2002generating,kern2016discoal},
a custom input file
format~\citep[e.g.][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
or an Application Programming Interface (API) to allow
models be defined programmatically~\citep[e.g.][]{
thornton2014cpp,thornton2019-nu,baumdicker2021-iu,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
Command line interfaces are a concise way of expressing
demographic models, and the syntax defined by \ms~\citep{hudson2002generating}
is used by a number of different
simulators~\citep[e.g.][]{ewing2010msms,chen2009fast,staab2015scrm},
and the POPDemog visualisation software~\citep{zhou2018popdemog}.
However, this conciseness means that models of even intermediate complexity
are difficult for humans to understand, making errors highly likely.
Input parameter file formats for simulators allow the model specification
to be less terse and (potentially) also allow for documentation in the
form of comments. However, there is no standard format shared across
simulators, and the details of the demographic models are not separated
from other simulation parameters. The APIs used by individual simulators
to define demographic models are also not appropriate as an interchange
standard because of the dependency on both the programming language involved,
and the specific details of the simulator in question.

Here we present ``Demes'', a data model and file format specification for
complex demographic models developed by the PopSim Consortium. The Demes data
model precisely defines the sizes of populations and the relationships
between populations in terms of migrations over time,
and provides a way to explicitly encode the
information relevant to demography while avoiding repetition. This data model
is implemented in the widely used YAML format~\citep{ben2009yaml}, which
provides a good balance between human and machine readability. The
specification precisely defines the required behaviour of implementations,
ensuring that there is no ambiguity of interpretation, and includes both a
reference implementation and an extensive suite of test examples and their
expected output. The initial software ecosystem includes high-quality Python
and C parser implementations, as well as utilities for verification and
visualisation of Demes models, and has been implemented in several popular
inference and simulation methods (Table~\ref{tab-software}).  We hope that this
data model and file format will be widely adopted by the community, such that
users can expect to simulate directly from inferred models with little or no
programming effort.

@@

The Demes specification is a formal data model for describing
the properties of populations over time,
along with some metadata and provenance information.
The data model is based on the ubiquitous JSON~\citep{bray2017javascript}
standard, and the structure formally defined using
JSON Schema~\citep{wright2020json}.
The Demes JSON schema rigorously defines the data model,
specifying the hierarchical structure and the properties that are allowed in different
contexts along with their types and permissible ranges. This schema, and its
accompanying documentation, fully describe the entities in the model and their
relationships and the required behaviour of implementations. Since the
schema is definitive, we will not recapitulate the details
here, but instead focus on the high level properties of the model and
the rationale behind key design decisions.

The design of Demes is a balance between two partially competing requirements:
that (a) models should be easy for humans to understand and manipulate; and (b)
software processing Demes models should be provided with an unambiguous
representation that is straightforward to process. For efficiency of
understanding and avoidance of model specification error, we require a data
representation without redundancy (i.e., repetition of values). But, for the
simplicity of software working with the Demes model (and the avoidance of
programming error, or divergence in interpretations of the specification) it is
preferable to have an explicit representation, in which all relevant values are
readily available. Thus, we consider three entities: the high-level, human
readable model specification; the low-level, fully-qualified model
specification; and the parser, which is responsible for transforming the former
into the latter.

\subsection*{Population genetics model}

@@

Population sizes are given as the number of individuals, and details
such as ploidy levels are considered external to the model.
We focus on the number of individuals as opposed to the number of genomes
Sizes and mating system details are specified for each population within
epochs.
Epochs are contiguous time intervals that define
the existence interval of the population. Each epoch specifies the population size
over that interval, which can be a constant value or function defined by start
and end sizes that must remain positive.

@@

\subsection*{High-level model specification}

@@

\subsection*{Parsers}
The implicit value inference and default value propagation required to
fully resolve the high-level Demes model description is straightforward
to describe, but still requires significant effort to implement in
software. Moreover, there are many constraints on the data model
(for example, epoch \texttt{end\_time} values should be strictly decreasing
within a deme), which must be enforced.
It would not be reasonable to require every program that
takes Demes models as input to implement this logic, as the programming
effort would be significant (limiting adoption)
and it is likely that if there were many implementations some would differ
in their details (harming the software ecosystem).
% GG: this is currently the case for programs accepting ms command lines. E.g.
%  - MaCS interprets -es proportion as 1-proportion compared to ms.
%  - SCRM uses different options for -ema command.
We therefore define
a standard software entity as part of the specification (the parser),
which performs this task, and which can be shared by programs that
support demes as an input. By having relatively few high-quality Demes
parsers available as libraries (ideally, one per programming language),
the probability of divergences from the standard is greatly reduced.

@@

\subsection*{Scope of the specification}
\subsubsection*{Static models, not parameterised models}
\jkcomment{This section is very rough.}

@@

\subsubsection*{Population-level features, not genome features}
Demes is designed to interface with a large number of simulation and inference
methods, so we have restricted to the specification to only describe
demographic features at the population level. As the underlying discrete
demographic structure is the common thread connecting many widely used
population genetic methods, many features of underlying genome biology are
omitted. Most notably, mutation and recombination rates are left unspecified
(though they can be encoded as metadata in the data structure). Genome ploidy
is left ambiguous, and selection and dominance models are absent. Using a Demes
model, which specifies deterministic population sizes, as the demographic input
for simulation using software that allows for selection implies that selection
is \emph{soft}, that is, genome fitnesses are rescaled relative to the mean
population fitness \citep{christiansen1975hard}. It is possible to instead
interpret population sizes as carrying capacities, allowing for a model of
\emph{hard} selection. However, this should not be understood as the default
interpretation of population sizes, which conventionally specifies a
deterministic number of individuals. Finally, as Demes assumes exchangeability
of individuals within populations, it doesn't provide a way to define geographic
spatial structure within populations.

Demes version 1.0 is intended to provide a stable and minimal specification
that is compatible with a wide range of simulators. To ensure consistency
across implementations, the specification is versioned and we commit to
following semantic versioning guidelines.

% Add a bit more space between rows.
\renewcommand{\arraystretch}{1.5}
\begin{table}
\begin{center}
\begin{tabular}{lp{12cm}}
\toprule
\multicolumn{2}{l}{Software infrastructure}\\
\midrule
\texttt{demes-python} &
    A Python library for loading, saving, and working with
    Demes models. Includes conversion utilities for existing
    demographic model descriptions such as
    \texttt{ms}~\citep{hudson2002generating}
    (\url{https://github.com/popsim-consortium/demes-python}).\\

\texttt{demes-c} &
    A C library for parsing Demes YAML descriptions.
    (\url{https://github.com/grahamgower/demes-c}). \\

\texttt{demesdraw} &
    A Python library for visualising Demes models
    (\url{https://github.com/grahamgower/demesdraw}). \\

\midrule
\multicolumn{2}{l}{Methods using Demes as input/output format}\\
\midrule

% TOOL AUTHORS: please add a row to this table to summarise what the software
% does, and how it implements demes.

\dadi &
    Summary of what dadi does and how it implements Demes~\citep{gutenkunst2009inferring}.
    \\

\fwdpy &
    Forward-time Wright-Fisher simulation \citep{thornton2014cpp,thornton2019-nu}\\

\gadma &
    Demographic inference \citep{noskova2020gadma}.\\

\moments &
    Compute expected diversity statistics (SFS, LD) and perform
    demographic inference using summary statistics
    \citep{jouganous2017inferring,ragsdale2019models}.\\

\msprime &
    Coalescent simulation of Demes models, including command line
    interface~\citep{kelleher2016efficient,kelleher2020coalescent,baumdicker2021-iu}.\\

\stdpopsim &
    A community-maintained collection of empirical data for species
    and demographic models specified in Demes
    format~\citep{adrion2020community}.\\

\bottomrule
\end{tabular}
\end{center}
\caption{\label{tab-software}
Software support for Demes. We have included software infrastructure developed
for working with Demes models (parsing, validation, visualisation, etc)
as well as all known downstream software that implements the specification,
at the time of writing.}
\end{table}

\bibliographystyle{plainnat}
\bibliography{paper}

\end{document}

>>>>

\usepackage{filecontents,catchfile}
\begin{filecontents*}{software-table.tex}
\end{filecontents*}

@@

\begin{abstract}
Understanding the demographic history of populations is a
key goal in population genetics, and with improving methods
and data, ever more complex models are being estimated.
Demographic models of current interest
typically consist of a set of discrete populations,
their sizes and growth rates, and continuous and pulse migrations
between those populations over a number of epochs, which can require
dozens of parameters to fully describe. There is currently
no standard format to define such models, significantly
hampering  progress in the field. In particular, the important
task of translating the model descriptions in published
work into input suitable for population genetic simulators is labour intensive
and error prone.
We propose the Demes data model and file format,
built on widely used technologies,
to alleviate these issues. Demes provides a well-defined and unambiguous
graphical model of populations and their properties that is straightforward to
implement in software, and a text file format that is designed for
simplicity and clarity.
We provide thoroughly tested implementations of Demes parsers in Python
and C, and showcase initial support in a number of simulators and inference
methods.
\end{abstract}

@@

The ever-increasing amount of genetic sequencing data from genetically and
geographically diverse species and populations has allowed us to infer complex
demography and study life history at fine scales.
An integral component to such population genetics studies is simulation.
Software to either simulate whole genome
sequences~\citep{thornton2014cpp,thornton2019-nu,staab2015scrm,baumdicker2021-iu,kelleher2016efficient,haller2019slim}
or informative summary statistics of
diversity~\citep{gutenkunst2009inferring,kamm2017efficient,jouganous2017inferring}
have enabled the increasing complexity of genomic studies, with several software
packages capable handling large sample sizes, many interacting populations, and
deviations from panmictic random-mating assumptions.
This ability to infer and simulate such complex demographic scenarios, however,
has highlighted a major shortcoming in community standards.
The fragmented landscape of different approaches to describing demographic
models makes it difficult to compare inferences made by different methods
and to reliably simulate from previously inferred models.

Inference methods for demography usually output a set of
text files which describe the inferred model,
and nearly all methods use different file formats.
Inference results are typically reported in publications
via a combination of visual depiction,
a list of key parameters in tabular form and a discussion within the text.
Unfortunately these descriptions are often ambiguous, and
implementing the precise model inferred for later simulation
is at best tedious and error prone~\citep{ragsdale2020lessons}
and occasionally impossible because of missing information.
%\aprcomment{This feels tangential - maybe discussion}
%The \stdpopsim\ project is a community-maintained catalog of species
%information and demographic models designed to provide standardised
%population genetic simulations~\citep{adrion2020community}.
%The development process for adding demographic models to
%\stdpopsim\ tries to avoid errors in model implementation by requiring two
%independent implementations of each model
%to be equivalent. While this process is
%robust and has uncovered both significant
%errors~\citep{ragsdale2020lessons} and ambiguities in
%published models, it is labour intensive.
Instead, it would be far better if published results were reported
in an unambiguous standardised format that could be provided
directly as input to simulators.

Simulation is a core tool in population genetics, and a
many methods have been developed over the past three
decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer}.
Simulations are based on highly idealised population models,
and one of the key uses of inferred demographic histories
is to make simulations more realistic.
Simulation methods take three broad approaches to specifying
the demographic model to simulate,
using either a command line
interface~\citep[e.g.][]{hudson2002generating,kern2016discoal},
a custom input file
format~\citep[e.g.][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
or an Application Programming Interface (API) to allow
models be defined programmatically~\citep[e.g.][]{
thornton2014cpp,thornton2019-nu,baumdicker2021-iu,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
Command line interfaces are a concise way of expressing
demographic models, and the syntax defined by \ms~\citep{hudson2002generating}
is used by a number of different
simulators~\citep[e.g.][]{ewing2010msms,chen2009fast,staab2015scrm},
and the POPDemog visualisation software~\citep{zhou2018popdemog}.
However, this conciseness means that models of even intermediate complexity
are difficult for humans to understand, making errors highly likely.
Input parameter file formats for simulators allow the model specification
to be less terse and allow for documentation in the form of comments.
However, there is no standard format shared across simulators,
and the details of the demographic models are not separated
from other simulation parameters.
The APIs used by individual simulators to define demographic models are also
not appropriate as an interchange standard because of the dependency on both
the programming language involved,
and the specific details of the simulator in question.

Here we present ``Demes'', a data model and file format specification for
complex demographic models developed by the
PopSim Consortium~\citep{adrion2020community}.
The Demes data model precisely defines the sizes and relationships
of populations, and it provides a way to explicitly encode the
information relevant to demography while avoiding repetition.
This data model is implemented in the widely used YAML
format~\citep{ben2009yaml}, which provides a good balance between human
and machine readability.
The specification precisely defines the required behaviour of implementations,
ensuring that there is no ambiguity of interpretation, and includes both a
reference implementation and an extensive suite of test examples and their
expected output.
The initial software ecosystem includes high-quality Python
and C parser implementations, as well as utilities for verification and
visualisation of Demes models, and has been implemented in several popular
inference and simulation methods (Table~\ref{tab-software}).
We hope that this data model and file format will be widely adopted
by the community, such that users can expect to simulate directly
from inferred models with little or no programming effort.

@@

The design of Demes is a balance between two partially competing requirements:
that (a) models should be easy for humans to understand and manipulate;
and (b) software processing Demes models should be provided with an unambiguous
representation that is straightforward to process.
For efficiency of understanding and avoidance of model specification error,
we require a data representation without redundancy (i.e., repetition of values).
However, for the simplicity of software working with the Demes model
(and the avoidance of programming error, or divergence in
interpretations of the specification) it is preferable to have an explicit
representation, in which all relevant values are readily available.
Thus, Demes is composed of three entities: the high-level, human-readable
model specification; the low-level, fully-qualified model
specification; and the parser, which is responsible for transforming the former
into the latter.

Below we provide a brief overview of the population genetics models
that Demes supports and the components of the Demes infrastructure.
We then highlight an example usage, including visualization and simulation
of a multi-population demographic model.
Detailed descriptions of the population genetic model, parsers, scope, and
additional examples are given in the Appendix.

\CatchFileDef{\softwaretable}{software-table.tex}{}
\renewcommand{\arraystretch}{1.5}
\begin{table}
\begin{center}
\begin{tabular}{lp{12cm}}
\softwaretable
\end{tabular}
\end{center}
\caption{\label{tab-software}
Software support for Demes. We have included software infrastructure developed
for working with Demes models (parsing, validation, visualisation, etc)
as well as all known downstream software that implements the specification,
at the time of writing.}
\end{table}

\subsection*{Population genetic model}
1 paragraph

\subsection*{High level specification}
1 paragraph

\subsection*{Parsers}
and languages. 1 paragraph

\subsection*{The scope of Demes}

\begin{itemize}
    \item Static, instead of parameterized models (1 paragraph)
    \item Population level instead of genomic features (1 paragraph)
\end{itemize}

\section*{Illustration: simulation using Demes}

\begin{itemize}
    \item Multi-population model with migrations and/or pulses
    \item Demesdraw
    \item msprime simulation
    \item moments simulation
    \item Compare SFS in each population
    \item The input YAML file should go in the Appendix
\end{itemize}

\bibliographystyle{genetics}
\bibliography{paper}

\renewcommand{\thefigure}{A\arabic{figure}}
\renewcommand{\thetable}{A\arabic{table}}
\renewcommand{\theequation}{A\arabic{equation}}
\renewcommand{\thesection}{A\arabic{section}}
\setcounter{figure}{0}
\setcounter{table}{0}
\setcounter{equation}{0}

\section*{Appendix}

The Demes specification is a formal data model for describing
the properties of populations over time,
along with some metadata and provenance information.
The data model is based on the ubiquitous JSON~\citep{bray2017javascript}
standard, and the structure is formally defined using
JSON Schema~\citep{wright2020json}.
The Demes JSON schema rigorously defines the data model,
specifying the hierarchical structure and the properties that are allowed in different
contexts along with their types and permissible ranges.
This schema, and its accompanying documentation,
fully describe the entities in the model and their
relationships and the required behaviour of implementations.
Since the schema is definitive, we will not recapitulate the details
here, but instead focus on the high level properties of the model and
the rationale behind key design decisions.

Below, we detail
\begin{itemize}
    \item The population genetics model in Section~\ref{sec:appendix-pop-gen-model}
    \item The high-level model specification in Section~\ref{sec:appendix-spec}
    \item The desired behaviour of Demes-format parsers in Section~\ref{sec:appendix-parsers}
    \item And the intended scope of Demes in Section~\ref{sec:appendix-scope}
\end{itemize}

\section{Population genetics model details}\label{sec:appendix-pop-gen-model}

@@

Population sizes are given as numbers of individuals, and details
such as ploidy levels are considered external to the model.
We therefore focus on the number of individuals as opposed to the number of genomes.
Sizes and mating system details are specified for each population within
epochs.
Epochs are contiguous time intervals that define
the existence interval of the population. Each epoch specifies the population size
over that interval, which can be a constant value or function defined by start
and end sizes that must remain positive.

@@

\section{High-level model specification}\label{sec:appendix-spec}

@@

\section{Parsers}\label{sec:appendix-parsers}

The implicit value inference and default value propagation required to
fully resolve the high-level Demes model description is straightforward
to describe, but still requires significant effort to implement in
software. Moreover, there are many constraints on the data model
(for example, epoch \texttt{end\_time} values should be strictly decreasing
within a deme), which must be enforced.
It would not be reasonable to require every program that
takes Demes models as input to implement this logic, as the programming
effort would be significant (limiting adoption)
and it is likely that if there were many implementations some would differ
in their details (harming the software ecosystem).
% GG: this is currently the case for programs accepting ms command lines. E.g.
%  - MaCS interprets -es proportion as 1-proportion compared to ms.
%  - SCRM uses different options for -ema command.
We therefore define
a standard software entity as part of the specification (the parser),
which performs this task, and which can be shared by programs that
support demes as an input. By having relatively few high-quality Demes
parsers available as libraries (ideally, one per programming language),
the probability of divergences from the standard is greatly reduced.

@@

\section{Scope of the specification}
\label{sec:appendix-scope}

\subsection{Static models, not parameterised models}
\label{sec:appendix-static}

\jkcomment{This section is very rough.}

@@

\subsection{Population-level features, not genome features}
\label{sec:appendix-features}

Demes is designed to interface with a large number of simulation and inference
methods, so we have restricted to the specification to only describe
demographic features at the population level. As the underlying discrete
demographic structure is the common thread connecting many widely used
population genetic methods, many features of underlying genome biology are
omitted. Most notably, mutation and recombination rates are left unspecified
(though they can be encoded as metadata in the data structure). Genome ploidy
is left ambiguous, and selection and dominance models are absent. Using a Demes
model, which specifies deterministic population sizes, as the demographic input
for simulation using software that allows for selection implies that selection
is \emph{soft}, that is, genome fitnesses are rescaled relative to the mean
population fitness \citep{christiansen1975hard}. It is possible to instead
interpret population sizes as carrying capacities, allowing for a model of
\emph{hard} selection. However, this should not be understood as the default
interpretation of population sizes, which conventionally specifies a
deterministic number of individuals. Finally, as Demes assumes exchangeability
of individuals within populations, it doesn't provide a way to define geographic
spatial structure within populations.

Demes version 1.0 is intended to provide a stable and minimal specification
that is compatible with a wide range of simulators. To ensure consistency
across implementations, the specification is versioned and we commit to
following semantic versioning guidelines.
\end{document}
